---
title: "setup_cmdstanr"
author: "Rich"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The aim of this file is to capture how to setup cmdstanr to run with brms.

The reason is that I arrived at ETHZ and R and R studio automatically updated via self-service and this caused problems for the toolchain and brms/stan. There appears to be an interaction between versions of R and R Studio and the operation of rstan. So much so that I am currently (as Ventura 13.3.1 (22E261) and the latest R Studio - 2023.03.1+446) not using rstan with brms (because it was problematic), but I am instead using cmdstanr with brms.

AFAIK, however, using cmdstanr might be better than using rstan anyway (see links in section 1 below). Cmdstanr is lightweight and is kept up-to-date more frequently compared to rstan, which is released via CRAN. So using cmdstan and cmdstanr seems like the way to go forward in the future.

The following is in two parts. 1) document the install process of cmdstanr. 2) document how to setup brms to use cmdstanr (instead of rstan).

## 1) document the install process of cmdstanr ##

some relevant links:

overview
https://mc-stan.org/cmdstanr/

getting started
https://mc-stan.org/cmdstanr/articles/cmdstanr.html

How does it work?
https://mc-stan.org/cmdstanr/articles/cmdstanr-internals.html

## install cmdstanr (and brms) ##

use github remotes

```{r, eval=FALSE}
install.packages("remotes")
remotes::install_github("stan-dev/cmdstanr")
```

and also install brms, as we will also use that later on

```{r, eval=FALSE}
install.packages("brms")
```

record the version in the renv lockfile

```{r, eval=FALSE}
renv::snapshot()
```

## CmdStanR requires a working installation of CmdStan on your local machine ##

to install cmdstan, run this chunk.

```{r, eval=FALSE}
install_cmdstan(cores=4) ## cores just an example and without a folder location
# specified it just uses the default folder location. The default is fine for most
#  setups though.

# and to specify the folder location, use the below...
# install_cmdstan(cores=4,
#                 dir = "/Users/Rich/.cmdstan")
```

then check the installation

```{r, eval=FALSE}
check_cmdstan_toolchain() 
# if successful, it should read...
# The C++ toolchain required for CmdStan is setup properly!
```

Note - of course, you can check the toolchain whenever to make sure things are ok and if you suspect they may be already setup, then you can run a quick check.

## As an aside, you can also manually download cmdstan ##

This is a non-essential sidepoint, but worth knowing about.

In some situations, using the install_cmdstan() function might not be possible and
in those situations you can just download cmdstan and place it on your machine. See here: https://discourse.mc-stan.org/t/how-to-install-cmdstanr-on-server-without-internet/26770

The steps are as follows:

- download from here: https://github.com/stan-dev/cmdstan/releases

- and save in the equivalent location: e.g., /Users/Rich/.cmdstan/cmdstan-2.32.2

- then run: cmdstanr::rebuild_cmdstan()

## load the package/s (like any other package) ##

e.g.,

```{r load-pkg}
pkg <- c("cmdstanr", "brms")

lapply(pkg, library, character.only = TRUE)
```

then modify settings in brms to use cmdstanr on the backend...

```{r}
options(brms.backend = "cmdstanr")
```
