---
title: "setup_cmdstanr"
author: "Rich"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The aim is to capture how to setup cmdstanr to run with brms.

The reason is that I arrived at ETHZ and R and R studio automatically updated via self-service and this caused problems for the toolchain and brms/stan. There appears to be an interaction between versions of R and R Studio and the operation of rstan. So much so that I am currently (as Ventura 13.3.1 (22E261) and the latest R Studio - 2023.03.1+446) not using rstan with brms (because it was problematic), but I am instead using cmdstanr with brms.

The following is in two parts. 1) document the install process of cmdstanr. 2) document how to setup brms to use cmdstanr (instead of rstan)

## 1) document the install process of cmdstanr ##

some relevant links:

overview
https://mc-stan.org/cmdstanr/

getting started
https://mc-stan.org/cmdstanr/articles/cmdstanr.html

How does it work?
https://mc-stan.org/cmdstanr/articles/cmdstanr-internals.html


I think this is what I did:

1. install
 we recommend running this is a fresh R session or restarting your current session
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))

e.g., 

```{r install-pkg}
# install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
# install.packages(c("brms", "here", "knitr", "tidyverse", "tidybayes", "bayesplot", "patchwork", "future")) # parallel comes pre-installed, hence it only needs loading in the next chunk and not installing.
```


update as of Aug 25, 2023

the above did not record cmdstanr in the renv.lock file or it did but it didn't know the repo / source.

instead install as follows:

use github remotes

```{r}
# install.packages("remotes")
remotes::install_github("stan-dev/cmdstanr")
```

However, it still requires cmdstan to be available on the machine and that is NOT in the lock file which is kind of weird?? So you still need to unpack cmdstan in the right place manually. 

only machines this has been:

Macbook
/Users/rramsey/.cmdstan/cmdstan-2.32.2

or

iMac
/Users/Rich/.cmdstan/cmdstan-2.32.2

And you download / install it as follows:

the solution is here: https://discourse.mc-stan.org/t/how-to-install-cmdstanr-on-server-without-internet/26770

- download from here: https://github.com/stan-dev/cmdstan/releases

- and save in the equivalent location: /Users/Rich/.cmdstan/cmdstan-2.32.2

- then run: cmdstanr::rebuild_cmdstan()






2. load (like any other package)

e.g.,

```{r load-pkg}
# pkg <- c("cmdstanr", "brms", "here", "knitr", "tidyverse", "tidybayes", "bayesplot", "patchwork", "future", "parallel")
# 
# lapply(pkg, library, character.only = TRUE)
```

3. CmdStanR requires a working installation of CmdStan

run this

```{r}
install_cmdstan(cores=4) ## cores just an example

# and to specifiy the folder location 
install_cmdstan(cores=4,
                dir = "/Users/Rich/.cmdstan")

rebuild_cmdstan(cores=4,
                dir = "/Users/Rich/.cmdstan")
```

then check it

```{r}
check_cmdstan_toolchain() 
# if successful, it should read...
# The C++ toolchain required for CmdStan is setup properly!
```

Note - of course, you can check the toolchain whenever to make sure things are ok and if you suspect they may be already setup, then you can run a quick check.

## 2) document how to setup brms to use cmdstanr (instead of rstan) ##

this is fairly straightforward (I think).

Once installed, load packages as usual. But remember to set options so that brms knows to use cmdstanr and not rstan because rstan is the default.

e.g., 

```{r}
# load packages as usual, e.g.,
# pkg <- c("cmdstanr", "brms")
# lapply(pkg, library, character.only = TRUE)

# then modify brms options to select cmdstanr
# options(brms.backend = "cmdstanr")
```
